### 2010-07-09 doc/guide/aggregates.tex
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3171 3e02f596-3131-0410-8352-e04eeaec37cc

\section{Input Languages}\label{sec:language}

This section provides an overview of the input languages of
grounder \gringo, combined grounder and solver \clingo,
incremental grounder and solver \iclingo, and of solver \clasp.
The joint input language of \gringo\ and \clingo\ is detailed in
Section~\ref{subsec:lang:gringo}.
It is extended by \iclingo\ with a few directives described in Section~\ref{subsec:lang:iclingo}.
Finally, Section~\ref{subsec:lang:clasp} is dedicated to the inputs
handled by \clasp.

\subsection{Input Language of \gringo\ and \clingo}\label{subsec:lang:gringo}

The tool \gringo~\cite{gescth07a} is a grounder capable of translating
logic programs provided by users into equivalent ground programs.
The output of \gringo\ can be piped into solver \clasp~\cite{gekanesc07b},
which then computes answer sets.
System \clingo\ internally couples \gringo\ and \clasp, thus,
it takes care of both grounding and solving.
% The input of \gringo\ and \clingo\ is the same,
% but 
In contrast to \gringo\ outputting ground programs, 
\clingo\ returns answer sets.
Both \gringo\ and \clingo\ can handle level-restricted input programs
(cf.\ Section~\ref{subsec:lambda}),
usually specified in one or more text files whose names are passed via the command line
in an invocation of either \gringo\ or \clingo.
We below provide a description of constructs belonging to the input language of \gringo\ and \clingo.

\subsubsection{Normal Programs and Integrity Constraints}\label{subsec:gringo:normal}

In the previous sections, we have already seen a number of normal programs.
Now also considering integrity constraints, we get the following basic rule types:

\begin{tabular}{rl@{}l}
\textbf{Fact:} & $A_0$&\code{.}
\\
\textbf{Rule:} & $A_0$& $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\\
\textbf{Integrity Constraint:} & & $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\end{tabular}
\\

\noindent
The head~$A_0$ of a fact or a rule is an \emph{atom} of the form
$p\code{(}t_1\code{,}\dots\code{,}t_m\code{)}$,
where $p$ is the name of some predicate, that is, a sequence of letters and digits
starting with a lowercase letter,
and any $t_i$ is a term.%
\footnote{An atom~$p$ without arguments is simply a sequence of letters and digits
  (starting with a lowercase letter).
  For such an atom~$p$, parentheses after the name are skipped, e.g.,
  \pred{p42X} could be an atom without arguments.
  Also note that lowercase letters include ``\code{\mus},''
  which can thus be used at any position within a predicate name.}
A \emph{term} $t$ starting with an uppercase letter followed by
a sequence of letters and digits (e.g., \var{X08x15}) is a variable name,
and integers are written as sequences of digits possibly preceded by ``\code{-}.''
In addition, a term can have the same syntactic structure as an atom
(e.g., \code{\pred{p}(\const{a},\const{1},\const{f}(\var{X}))}
 can be either an atom or a term, depending on where it occurs in a rule),
and functions can be nested within a term.
There are also built-in constructs
(cf.\ Section~\ref{subsec:gringo:arith} and~\ref{subsec:gringo:comp})
having particular representations.
Finally, any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
for an atom~$A$.

In Section~\ref{subsec:semantics},
we have already provided a translation~$\psi$
from the head of a rule to a propositional formula.
By viewing an integrity constraint
$\code{:-}~\mathit{Body}\code{.}$ as a shorthand for $\bot~\code{:-}~\mathit{Body}\code{.}$,
we can now explain the semantics of integrity constraints by adding the following case:
\begin{itemize}
\item $\psi[\bot]=\bot$.
\end{itemize}
Note the role of integrity constraints is to eliminate answer set candidates.
In fact, given an integrity constraint~$\code{:-}~\mathit{Body}\code{.}$,
any answer set~$X$ is such that $X\not\models\phi[\mathit{Body}]$, or
in other words, some literal~$L$ in $\mathit{Body}$ must be unsatisfied w.r.t.~$X$.
Elaborate examples on the usage of facts, rules, and integrity constraints
will be provided in Section~\ref{sec:examples}.

\subsubsection{Classical Negation}\label{subsec:gringo:negation}

In logic programs, connective \code{not} expresses default negation,
that is, a literal $\code{not}~A$ is assumed to hold unless~$A$ is derived.
In contrast, the classical (or strong) negation of some proposition
holds if the complement of the proposition is objectively derived~\cite{gellif91a}.
Classical negation, indicated by symbol ``\code{-},'' is permitted in front of atoms.
That is, if~$A$ is an atom, then~$\code{-}A$ is the complement of~$A$.
Semantically, $\code{-}A$ is simply a new atom,
with the additional condition that~$A$ and~$\code{-}A$ must not jointly hold.
A logic program~$\Pi$ containing complementary atoms is thus translated
to propositional theory
$\Phi[\Pi \cup \{\code{:-}~A\code{,}\code{-}A \mid A\in\ground{\mathcal{A}}\}]$.%
\footnote{Recall that $\ground{\mathcal{A}}$ consists of all variable-free atoms
built from predicates and functions in~$\Pi$.}
Observe that classical negation is merely a syntactic feature that can be
implemented via integrity constraints whose effect is to eliminate
any answer set candidate containing complementary atoms.

\begin{example}\label{ex:flies:neg}
The following logic program reformulates the one in Example~\ref{ex:flies}:
%
\lstinputlisting[nolol]{examples/bird.lp}
\lstinputlisting[firstnumber=3]{examples/flycn.lp}
%
Logically, classical negation is reflected by
(implicit) integrity constraints as follows:%
\marginlabel{By invoking\\
  \code{\mbox{~}gringo -t \textbackslash\\
        \mbox{~}examples/bird.lp \textbackslash\\
        \mbox{~}examples/flycn.lp}
the reader can observe that \gringo\ indeed produces the integrity constraint
in Line~7.}
%
\begin{lstlisting}[firstnumber=6]
:- flies(tux),    -flies(tux).
:- flies(tweety), -flies(tweety).
\end{lstlisting}
The additional integrity constraints do not yet change the semantics of the
program, which still has the answer sets already provided in Example~\ref{ex:flies:ground}
(of course, identifying
\code{\pred{neg\mus flies}(\const{tux})} with
\code{\pred{-flies}(\const{tux})} and
\code{\pred{neg\mus flies}(\const{tweety})} with
\code{\pred{-flies}(\const{tweety})}).
This situation changes if we add the following fact:
\begin{lstlisting}[firstnumber=8]
flies(tux).
\end{lstlisting}
While the program from Example~\ref{ex:flies} still admits two answer sets,
both containing 
\code{\pred{flies}(\const{tux})} and
\code{\pred{neg\mus flies}(\const{tux})},
there no longer is any answer set for our new program using classical negation.
In fact, answer set candidates that contain both
\code{\pred{flies}(\const{tux})} and
\code{\pred{-flies}(\const{tux})} violate the integrity constraint in Line~6.
\eexample
\end{example}

\subsubsection{Disjunction}\label{subsec:gringo:disjunction}

Disjunctive logic programs permit connective~``\code{|}'' between atoms in rule heads.
An additional case of translation~$\psi$ from Section~\ref{subsec:semantics}
reflects the semantics of disjunction:
\begin{itemize}
\item $\psi[G\code{|}H]=\psi[G]\vee\psi[H]$.
\end{itemize}
%
The concept of level-restrictedness (cf.\ Section~\ref{subsec:lambda}) is extended
to disjunctive programs by,
for a disjunctive rule~$r$, every predicate~$p/i$
of some atom in the head of~$r$, and each $V\in\vars{r}$,
requiring that there is some $A\in\binders{V}{r}$ such that
$\lambda(p/i)>\lambda(q/j)$ for predicate~$q/j$ of~$A$.
Furthermore, a (maximal) stratified subprogram (cf.\ Section~\ref{subsec:strat})
must not contain any rule with a (non-trivial) disjunctive head.

The following rule combines the ones in Line~3 and~4 from Example~\ref{ex:flies:neg}:
\begin{lstlisting}[numbers=none]
flies(X) | -flies(X) :- bird(X).
\end{lstlisting}
In general, the use of disjunction however increases
computational complexity~\cite{eitgot95a}.
This is why \clingo%
\footnote{Run as a monolithic system performing both grounding and solving.}
and solvers like 
\assat~\cite{linzha04a},
\clasp~\cite{gekanesc07b},
\nomorepp~\cite{angelinesc05c},
\smodels~\cite{siniso02a}, and
\smodelscc~\cite{warsch04a}
do not work on disjunctive programs.
Rather,
\claspD~\cite{drgegrkakoossc08a},
\cmodels~\cite{gilima06a,lierler05a}, or
\gnt~\cite{janisesiyo06a}
needs to be used for solving a disjunctive program.%
\footnote{System \dlv~\cite{dlv03a} also deals with disjunctive programs,
  but it uses a different syntax than presented here.}
We thus suggest to use ``choice constructs'' (cf.\ Section~\ref{subsec:gringo:aggregate})
instead of disjunction, unless the latter is required for complexity reasons
(see~\cite{eitpol06a} for an implementation methodology in disjunctive ASP).

\subsubsection{Built-In Arithmetic Functions}\label{subsec:gringo:arith}

\gringo\ and \clingo\ support a number of arithmetic functions that
are evaluated during grounding.
The following symbols are used for these functions:
\code{+} (addition),
\code{-} (subtraction),
\code{*} (multiplication),
\code{/} or \code{div} (integer division),
\code{mod} (modulo function),
\code{abs} (absolute value),
\code{\textasciitilde} (bitwise complement),
\code{\&}  (bitwise AND),
\code{?}   (bitwise OR), and
\code{\^} (bitwise exclusive OR).

\begin{example}\label{ex:arith:fun}
The usage of arithmetic functions is illustrated by the logic program:%
\marginlabel{%
  The unique answer set of the program,
  obtained after evaluating all arithmetic functions,
  can be inspected by invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/arithf.lp}}
%
\lstinputlisting{examples/arithf.lp}
%
Note that variables~\var{L} and~\var{R} are instantiated to~\const{7} and~\const{2},
respectively, before arithmetic evaluations.
Consecutive and non-separative (e.g., before ``\code{(}'')
spaces can also be dropped,
while spaces around tokens \code{div} and \code{mod} are mandatory.
Furthermore, the argument of function \code{abs} must be enclosed in parentheses.
In Line~9, observe that there is a unary version of \code{-},
``\code{- \var{R}}'' standing for ``\code{0 - \var{R}}.''
The four bitwise functions apply to signed integers,
using the complement on two of a negative integer.
\eexample
\end{example}

It is important to note that variables
in the scope of an arithmetic function are not necessarily bound 
(in the sense of Section~\ref{subsec:lambda}) by a corresponding atom.
For instance, atom \code{\pred{p}(\var{X}+1,\var{X})} belongs to
$\binders{\var{X}}{r}$, while atom \code{\pred{p}(\var{X}+1,\var{Y})}
does not belong to $\binders{\var{X}}{r}$ because it contains~\var{X}
only in the scope of an arithmetic function.

\subsubsection{Built-In Comparison Predicates}\label{subsec:gringo:comp}

The following built-in predicates permit term comparisons
within the bodies of rules and on the right-hand side of conditions
(cf.\ Section~\ref{subsec:gringo:condition}):
\code{==} (equal),
\code{!=} (not equal),
\code{<} (less than),
\code{<=} (less than or equal),
\code{>} (greater than),
\code{>=} (greater than or equal).

\begin{example}\label{ex:arith:pred}
The usage of comparison predicates is illustrated by the logic program:%
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/arithc.lp}}
%
\lstinputlisting{examples/arithc.lp}
%
The last two lines hint at the fact that arithmetic functions are evaluated
before comparison predicates, so that the latter actually compare integers.

All comparison predicates can also be used with constants,
as in the next program:%
\marginlabel{%
  As above, invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/symbc.lp}
  yields the unique answer set of the program in terms of facts.}
%
\lstinputlisting{examples/symbc.lp}
%
Finally, note that~\code{==} and~\code{!=} also apply to
compound terms over functions with non-zero arity as well as to
mixed integer and non-integer arguments.%
\footnote{%
  Such functionalities are currently not supported for the other four
  comparison predicates,
  where (after instantiation and arithmetic evaluation)
  both arguments must be either integers or (symbolic) constants.
}
\eexample
\end{example}

Importantly, a built-in comparison predicate cannot bind variables,
that is, it does not belong to $\binders{\var{X}}{r}$ for any variable~\var{X}.
Also note that built-in predicates are not considered by level mappings~$\xi$
in the context of stratification (cf.\ Section~\ref{subsec:strat}), or equivalently,
one may assume that built-in predicates are mapped to level~$0$ and all other
predicates to levels greater than~$0$.

\subsubsection{Assignments}\label{subsec:gringo:assign}

Built-in predicate~\code{=} can be used in the body of a rule
(or on right-hand sides of conditions introduced in Section~\ref{subsec:gringo:condition})
to assign a term on its right-hand side to a variable on its left-hand side.
Note that ``\code{\var{X} \!=\! }$t$'' belongs to~$\binders{\var{X}}{r}$,
provided that~$t$ is variable-free or that its variables are bound
by atoms other than assignments with~\var{X} on the right-hand side.
Cyclic assignments over otherwise unbound variables are thus excluded.

\begin{example}\label{ex:assign}
The next program demonstrates how terms can be assigned to variables:%
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/assign.lp}}
%
\lstinputlisting{examples/assign.lp}
%
Line~3 contains four assignments, where the right-hand sides directly or indirectly
depend on~\var{X} and~\var{Y}.
These two variables are bound in Line~4 via atoms of predicate \pred{num}/$1$.
Also observe the different usage and role of built-in comparison predicate~\code{==}.
\eexample
\end{example}

\subsubsection{Intervals}\label{subsec:gringo:interval}

In Line~1 of Example~\ref{ex:assign},
there are five facts~\code{\pred{num}($k$)}
over consecutive integers~$k$.
For a more compact representation,
\gringo\ and \clingo\ support integer intervals of the form $i$\code{..}$j$,
where~$i$ and~$j$ are integers.
Such an interval represents each integer~$k$ such that $i\leq k\leq j$,
and intervals are expanded during grounding.

\begin{example}\label{ex:int}
The next program makes use of integer intervals:
%
\lstinputlisting{examples/int.lp}
%
The facts in Line~1 and~2 are expanded as follows:
%
\begin{lstlisting}[numbers=none]
num(1).    num(2).   num(3).   num(4).   num(5).
top5(5).  top5(6).  top5(7).  top5(8).  top5(9).
\end{lstlisting}
%
By instantiating~\var{X} to~\const{9},
the rule in Line~4 becomes:
%
\begin{lstlisting}[numbers=none]
top5num(1..5,5..9) :- num(5..9), top5(1..5), top(9).
\end{lstlisting}
%
It is expanded to the cross product
$(\code{\const{1}..\const{5}})\times
 (\code{\const{5}..\const{9}})\times
 (\code{\const{5}..\const{9}})\times
 (\code{\const{1}..\const{5}})$
of intervals:
%
\begin{lstlisting}[numbers=none,escapechar=@]
top5num(1,5) :- num(5), top5(1), top(9).
top5num(2,5) :- num(5), top5(1), top(9).
        @\rlap{\vdots}@
top5num(5,5) :- num(5), top5(1), top(9).
top5num(1,6) :- num(5), top5(1), top(9).
top5num(2,6) :- num(5), top5(1), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@
top5num(5,9) :- num(5), top5(1), top(9).
top5num(1,5) :- num(6), top5(1), top(9).
top5num(2,5) :- num(6), top5(1), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@          @\rlap{\vdots}@
top5num(5,9) :- num(9), top5(1), top(9).
top5num(1,5) :- num(5), top5(2), top(9).
top5num(2,5) :- num(5), top5(2), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@          @\rlap{\vdots}@         @\rlap{\vdots}@
top5num(5,9) :- num(9), top5(4), top(9).
top5num(1,5) :- num(5), top5(5), top(9).
top5num(2,5) :- num(5), top5(5), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@
top5num(5,9) :- num(5), top5(5), top(9).
top5num(1,5) :- num(6), top5(5), top(9).
top5num(2,5) :- num(6), top5(5), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@          @\rlap{\vdots}@
top5num(5,9) :- num(9), top5(5), top(9).
\end{lstlisting}
%
Note that only the rules with
\code{\pred{num}(\const{5})} and
\code{\pred{top5}(\const{5})}
\marginlabel{%
  Again the unique answer set is obtained via call:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/int.lp}}
%
in the body actually contribute to the unique answer set
of the above program by deriving all atoms \code{\pred{top5num}($m$,$n$)}
for $\const{1}\leq m\leq\const{5}$ and $\const{5}\leq n\leq\const{9}$.
\eexample
\end{example}

As with built-in arithmetic functions,
an integer interval mentioning some variable (like~\var{X} in Line~4 of Example~\ref{ex:int})
cannot be used to bind the variable.

\subsubsection{Conditions}\label{subsec:gringo:condition}

Conditions allow for instantiating variables to
collections of terms within a single rule.
This is particularly useful for encoding conjunctions or disjunctions over
arbitrarily many ground atoms as well as for the compact representation of aggregates
(cf.\ Section~\ref{subsec:gringo:aggregate}).
The symbol~``\code{:}'' is used to formulate conditions.

>>>>

\com{Rewrote much here}

An aggregate is an operation on a multiset of weighted literals that evaluates to a value,
depending on the type of aggregate this might be integers but, 
e.g., \const{\#min} and \const{\#max} aggregates also permit function symbols.
In combination with comparisons, we can extract a truth value from
an aggregate's evaluation, thus, obtaining an aggregate atom.
Customizing the notation in~\cite{ferraris05a},
we consider ground \emph{aggregate atoms} of the form:
%
\begin{equation}\label{eq:aggregate:atom}
  l~ 
  \mathrm{op} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  u
\end{equation}
%
where~$l$ and~$u$ are values each of which can possibly be omitted,
$\mathrm{op}$ is a function from multisets of values to a value,
each $L_i$ is a literal of the form~$A$ or $\code{not}~A$
for some atom~$A\in\ground{\mathcal{A}}$,
and each~$w_i$ is a value.
The intuitive reading of an aggregate atom is that the value returned by~$\mathrm{op}$
applied to all weights~$w_i$ such that~$L_i$ holds should be in-between
lower bound~$l$ and upper bound~$u$ (inclusively).
If~$l$ or~$u$ is omitted,
it means that the aggregate's value is not constrained from below or above,
respectively.

For an aggregate atom~$\mathit{Agg}$ as in Equation(\ref{eq:aggregate:atom}),
we let $\atom{\mathit{Agg}}=\{L_i\in\nolinebreak\ground{\mathcal{A}}\mid 1\leq i\leq n\}$
denote the set of atoms occurring positively in~$\mathit{Agg}$.
Currently, \gringo\ and \clingo\ support aggregate operations
\const{\#sum}, \const{\#min}, and \const{\#max}, \const{\#avg}, \const{\#even}, and \const{\#odd}.%
\footnote{%
  The additional operation \const{count} is a shorthand for a \const{sum}
  aggregate in which all weights are~\const{1}.}
We can now extend translations~$\phi$ and~$\psi$ from Section~\ref{subsec:semantics}
to aggregate atoms, 
using the approach in~\cite{ferraris05a} but
applying the ``choice semantics'' for aggregates
in heads of rules~\cite{ferlif05a,siniso02a}:%

\newcommand\shiftleft{\kern\leftshift&\kern-\leftshift}

\newlength{\leftshift} 
\setlength{\leftshift} {2em}
%\settowidth{\leftshift}{$\phi^{+}[ l~ \const{\#sum} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ]$}

\begin{align}
  % l op u (head) %
  \shiftleft
  \psi[ \lhd~ \mathrm{op} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]}\rhd ] 
  \label{eq:aggr_head} \\
   &= \phi[ \triangleleft~ \mathrm{op} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]}\triangleright ] \notag \\
   &\wedge \bigl(\bigwedge\nolimits_{A\in atom(\triangleleft~ \mathrm{op} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]}\triangleright)}(A\vee\neg A)\bigr) \notag \\
  % l op u %
  \shiftleft
  \phi[ l~ \mathrm{op} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} u ]
  \label{eq:aggr_lu} \\
  &= \phi[ l~ \mathrm{op} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ] \wedge{} 
    \phi[ \mathrm{op} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} u ] \notag \\
  % op %
  \shiftleft
  \phi[ \mathrm{op} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ] = \top
  \label{eq:aggr}
\end{align}
Translation~$\psi$ of an aggregate atom in the head of a rule consists of two parts:
first, $\phi$ is used as defined in the following
(in Equation (\ref{eq:aggr_head}), $\lhd$ and~$\rhd$ are used as ``wildcards'' for provided or omitted 
 lower and upper bounds),
second, a disjunction $(A\vee\neg A)$ is added for each atom~$A$ occurring positively.
Even though such disjunctions are tautological,
they are important under answer set semantics (cf.\ Section~\ref{subsec:semantics}),
as they permit deriving an arbitrary subset of atoms from an aggregate in the head
(which explains the name  ``choice semantics'').
Equation \ref{eq:aggr_lu} splits up an aggregate with lower and an upper into the conjunction of two
conjunctively connected parts.
Note that this split agrees with the semtaics in \cite{ferraris05a} independend of the aggregate $\mathrm{op}$ used.
Finally, Equation (\ref{eq:aggr}) reflects that an unconstrained aggregate atom always holds.

\paragraph{Sum Aggregates}
First, we show how monotone aggregates can be translated into a nested expression.
Using this translation we show how to transfom antimonotone sum aggregates and 
also give a translation that treats negative weights.
\footnote{%
  The translation~$\phi$ we provide for \const{sum} aggregates
  assumes that there are no negative weights for literals,
  while the more sophisticated translation in~\cite{ferraris05a} also
  captures negative weights.
  All current solvers work with positive weights only, and thus
  \gringo\ and \lparse\ use a compilation technique~\cite{siniso02a}
  to eliminate negative weights, which are not permitted in
  \lparse's output format~\cite{lparseManual}.
  As this technique may lead to counterintuitive results~\cite{ferraris05a},
  we strongly recommend not to use negative weights within \const{sum} aggregates.}
\begin{align}
  % l sum+ %
  \shiftleft
  \phi[ l~ \const{\#sum}^{+} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ] 
  \label{eq:sum_plus}\\
  &= \bigvee\nolimits_{\{i_1,\dots,i_j\}\subseteq\{1,\dots,n\},l\leq w_{i_1}+\dots+w_{i_j}}
   (\phi[L_{i_1}]\wedge\dots\wedge \phi[L_{i_j}]) \notag 
\end{align}
Equation (\ref{eq:sum_plus}) translates a monotone aggregate~$\const{\#sum}^{+}$
with a lower bound and only positive weights into a nested expressions.

\begin{align}
  % l sum %
  \shiftleft
  \phi[l~\const{\#sum}~\code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ]
  \label{eq:sum_neg} \\
  &= \phi [ l + \phi^{+}(w_1) + \dots + \phi^{+}(w_n) ~\const{\#sum}^{+}
    \code{[} \phi^{+}(L_1\code{=}w_1)\code{,} \dots\code{,} \phi^{+}(L_n\code{=}w_n) \code{]} ] \notag \\
  % weight lit %
  \shiftleft
  \phi^{+}(L=w) = 
    \left\{\begin{aligned}
      L=w & \mbox{ if } w \geq 0 \\
      \overline{L}=-w &  \mbox{ otherwise}\\
    \end{aligned}\right. \notag \\
  % weight %
  \shiftleft
  \phi^{+}(w) =-\min\{0,w\} \notag
\end{align}
Equation (\ref{eq:sum_neg}) removes all negative weights from a sum aggregate with a lower bound 
and then applies the transformation described above.
Whenever there is a negative weight, the sign of both literal and weight is inverted and the weight itself is subtracted from the bound.
Note that this transformation produces counter-intuitive results as soon as the aggregate is used recursively.
Nevertheless, there are scenarios where negative weights might be helpfull.
As long as the aggregate is stratified, the transformation corresponds with the intuitive meaning, e.g.,
you can use negative weights to compare two sets $A=\{a,b,c\}$ and $B=\{d,e\}$:
\begin{lstlisting}[numbers=none]
subseteq(A,B) :- #sum [ a=1,b=1,c=1,d=-1,e=-1 ] ] 0.
\end{lstlisting}

\begin{align}
  % sum u %
  \phi[ \const{\#sum} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]}~u ] 
  &= \neg \phi[  u+1~\const{\#sum} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]}]
  \label{eq:sum_u}
\end{align}
Finally, Equation (\ref{eq:sum_u}) translates sum aggregates with an upper bound using the
translation for sum aggregates with lower bounds described above.
Note that again negative literals might cause problems.
Even in the case that all weights are negative (and as such the sum aggregate itself is monotone),
the above translation is used, which might lead to counter-intuitive results.

\paragraph{Max and Min Aggregates} 
We define the sematics of max aggregates using the transformation for sum aggregates described above.
First, we translate max aggregates with upper and lower bound, respectively, into corresponding sum aggregates.
Then we give the semantics of min aggregates using the translation for max aggregates.
Note that in constrast to sum aggregates, max aggregates allow for using non-integer weights as long as there
is some strict total ordering defined on them.
\begin{align}
  % l max
  \shiftleft
  \phi[l~\const{\#max} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ]
  \label{eq:max_l}\\
  &= \phi[1~\const{\#sum} \code{[} L_1\code{=}\mathrm{ge}(w_1, l)\code{,} \dots\code{,} L_n\code{=}\mathrm{ge}(w_1, l) \code{]} ]  \notag \\
  % max(w,l)
  \shiftleft
  \mathrm{ge}(a, b) = \left\{\!\begin{aligned}
       1 &\mbox{ if } a \geq b \\
       0 &\mbox{ otherwise}
     \end{aligned}\right. \notag
\end{align}
A (monotone) max aggregate with a lower bound is translated into a sum by setting all weights 
that are greater equal to the lower bound to one and requiring that the sum of weights is at least one.
This transformation preserves the monotonicity of the aggregate and holds independendly of the type of weights used.

\begin{align}
  % max u %
  \shiftleft
  \phi[\const{\#max} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} u]
  \label{eq:max_u} \\
  &= \phi[\const{\#sum} \code{[} L_1\code{=}\mathrm{ge}(w_1, l)\code{,} \dots\code{,} L_n\code{=}\mathrm{ge}(w_1, l) \code{]} 0]  \notag \\
  \shiftleft
  \mathrm{gt}(a, b) = \left\{\!\begin{aligned}
       1 &\mbox{ if } a > b \\
       0 &\mbox{ otherwise}
     \end{aligned}\right. \notag
\end{align}
Similarly to Equation~(\ref{eq:max_l}), an (antimonotone) max aggregate with an upper bound can be translated into a sum with an upper bound.
Again the monotonicity of the aggregate is preserved.

\begin{align}
  % min u %
  \shiftleft
  \phi[ \const{\#min} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]}~u ] 
  \label{eq:min_u}  \\
  &= \phi[1~\const{\#sum} \code{[} L_1\code{=}\mathrm{ge}(l,w_1)\code{,} \dots\code{,} L_n\code{=}\mathrm{ge}(l,w_1) \code{]}]
  \notag \\
  % l min %
  \shiftleft
  \phi[ l~ \const{\#min} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ]
  \label{eq:min_l} \\
  &= \phi[\const{\#sum} \code{[} L_1\code{=}\mathrm{gt}(l,w_1)\code{,} \dots\code{,} L_n\code{=}\mathrm{gt}(l,w_1) \code{]} 0]
  \notag
\end{align}
Analogous to the definition of max aggregates, 
Equations (\ref{eq:min_l}) and (\ref{eq:min_u}) define the semantics of min aggregates.

\paragraph{Average Aggregates}
Avergate can be directly translated into sum aggregates by modifying the weights of its literals.
To see how the translation works consider:
\begin{align*}
\mathrm{avg}\code{[}a_1,\dots,a_n\code{]} &\prec b \\
a_1+\dots+a_n               &\prec n \cdot b \\
a_1 - b + \dots + a_n - b   &\prec 0
\end{align*}
where $\prec$ is some relation.
To avoid problems with empty average aggregates we define that $\mathrm{avg}\code{[} \code{]} \prec b$ iff $0 \prec 0$.
Thus the translation can turn an average aggregate into a sum by subtracting the bound from all its weights.
A strong point of this translation is that it preserves the monotonicity of the average aggregate.
This translation is applied in Equations (\ref{eq:avg_l}) and (\ref{eq:avg_u}).

\begin{align}
  % l avg %
  \shiftleft
  \phi[ l~\const{\#avg} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ] 
  \label{eq:avg_l} \\
  &= \phi[ \const{\#sum} \code{[} L_1\code{=}w_1-l \code{,} \dots\code{,} L_n\code{=}w_n-l \code{]}~0 ] 
  \notag \\
  % avg u %
  \shiftleft
  \phi[ \const{\#avg} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]}~u ] = 
  \label{eq:avg_u} \\
  &= \phi[ \const{\#sum} \code{[} L_1\code{=}w_1-u \code{,} \dots\code{,} L_n\code{=}w_n-u \code{]}~0 ]
  \notag
\end{align}

\paragraph{Parity Aggregates}
We also support the two parity aggregates $\code{\#even}$ and $\code{\#odd}$.
Intuitively, an even or odd aggregate is true if the sum of its weights modulo two is one or zero, respectively.
Note that we do not write bounds
Even and odd aggregates are transformed into nested expressions:
\begin{align}
  % odd %
  \shiftleft
  \phi[ \const{\#odd} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ] \\
  &=\left\{\begin{aligned}
    \bot & \mbox{ if } n = 0 \\
    \phi[ \const{\#odd} \code{[} L_2\code{=}w_2-u \code{,} \dots\code{,} L_n\code{=}w_n-u \code{]}~0 ] & \mbox{ if } 2 | w_1  \\
    \neg L_1 \wedge \phi[ \const{\#odd} \code{[} L_2\code{=}w_2 \code{,} \dots\code{,} L_n\code{=}w_n \code{]}~0 ] & \\
    \vee L_1 \wedge \neg \phi[ \const{\#odd} \code{[} L_2\code{=}w_2 \code{,} \dots\code{,} L_n\code{=}w_n \code{]}~0 ] & \mbox{ otherwise}
  \end{aligned}\right. \notag
\end{align}

\begin{align}
  % even %
  \shiftleft
  \phi[ \const{\#even} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ] \\
  &=\left\{\begin{aligned}
    \top & \mbox{ if } n = 0 \\
    \phi[ \const{\#even} \code{[} L_2\code{=}w_2-u \code{,} \dots\code{,} L_n\code{=}w_n-u \code{]}~0 ] & \mbox{ if } 2 | w_1  \\
    \neg L_1 \wedge \phi[ \const{\#even} \code{[} L_2\code{=}w_2 \code{,} \dots\code{,} L_n\code{=}w_n \code{]}~0 ] & \\
    \vee L_1 \wedge \neg \phi[ \const{\#even} \code{[} L_2\code{=}w_2 \code{,} \dots\code{,} L_n\code{=}w_n \code{]}~0 ] & \mbox{ otherwise}
  \end{aligned}\right. \notag
\end{align}

Item~3--8 specify these parts individually for~\const{sum},~\const{max}, and~\const{min}
aggregates, respectively.
Provided that only non-negative weights are used within \const{sum}s,
all three aggregates are subject to one comparison that is monotone,
meaning that by default it evaluates to \textit{false} so that truth must be established.
For \const{sum} and \const{max} aggregates, these monotone comparisons are related
to lower bounds (Item~3 and~5),
while an upper bound behaves monotonically for \const{min} (Item~8).
In turn, the opposite comparison is antimonotone, that is,
by default it evaluates to \textit{true} unless too many or improper literals hold.
This is the case for comparisons to upper bounds with aggregates
\const{sum} and \const{max} (Item~4 and~6), while it concerns 
the lower bound for \const{min} (Item~7).
% <cut here>
Beyond this special treatment in rule heads,
the truth values of aggregate atoms result from their literals as
may be intuitively expected,
and Item~1--8 basically do nothing but formalizing the standard meanings
of aggregate operations.
Note that the above translations merely define the semantics of aggregates,
which does not imply that solvers unfold them in this way.
In fact, \clasp\ natively supports aggregates without translating them.
